Carrier phase and amplitude estimation for phase shift keying with unequal error protection

### 2012-12-04 other/uep/paper.tex
added to uep document.

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-phase shift keying, and consider specifically the case where a number of different constellations sizes $M$ are used simultaneously.  We describe an algorithm to compute the least squares estimators of carrier phase and amplitude that requires only $O(L \log L)$ arithmetic operations, where $L$ is the number of recieved symbols.   
\end{abstract}
\begin{IEEEkeywords}
Coherent detection, noncoherent detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 1$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e. $P \cap D = \emptyset$, and $L = \abs{P \cup D}$ is the total number of symbols transmitted.  The data symbols are further separated into subsets according to the size of the constellation used to modulate each symbol.  Let $D_2,D_3,\dots$ denote a partition of the indices in $D$ such that $D_2$ is the set of indices corresponding to data symbols modulated with $2$-PSK (i.e BSPK), and $D_3$ is the set of symbols modulated with $3$-PSK, and $D_4$ is the set of symbols modulated with $4$-PSK (i.e. QPSK) and so on.  Let $G$ be the set of integers for which $D_m$ is not empty, i.e. $D_m \neq \emptyset$ whenever $m \in G$.  We have 
\[
D = \bigcup_{m=2}^{\infty}D_m = \bigcup_{m\in G} D_m.
\]

@@

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set) and where only one constellation size is used, i.e., $G = \{M\}$ contains precisely one element.  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the recieved signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  It is not entirely obvious how pilot symbols or data symbols from mulitple constellations should be included in this estimator.  One approach is to choose,
\[
A = \frac{1}{L} \sum_{i \in P} F(\abs{y_i}) \tfrac{y_i}{y_i} p_i^* +  \frac{1}{L} \sum_{m \in G} \sum_{i \in D_m} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^m.
\]
The simulations we present in Section~\ref{sec:simulations} suggest that the accuracy of this estimator is poor when compared to the least squares estimator.  %This problem does not occur with the least square estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Swelden's considered only the noncoherent setting and where only one constellation size is used, but we show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols and multiple constellation sizes. Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*, \label{eq:SS}
\end{align}
where $*$ denotes the complex congugate.  Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find the least squares estimator of $a_0$ as a function of $\{d_i, i \in D\}$,
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain the $SS$ conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  It turns out that there are atmost $(M+1)\abs{D}$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

We only require to consider $\theta$ in the interval $[0, 2\pi)$.  Consider how $\hat{d}_i(\theta)$ changes as $\theta$ varies from $0$ to $2\pi$.  Let $b_i = \hat{d}_i(0)$ and let 
\[
z_i = \angle{y_i} - \hat{u}_i(0) = \angle{y_i} - \round{\angle{y_i}}_{m}.
\]
Then,
\begin{equation}\label{eq:uicombos}
\hat{d}_i(\theta) = 
\begin{cases}
b_i, &  0 \leq \theta < z_i + \frac{\pi}{m} \\
b_i e^{-j2\pi/m}, & z_i + \frac{\pi}{m} \leq \theta < z_i + \frac{3\pi}{m} \\ 
\vdots & \\
b_i e^{-j2\pi k /m}, & z_i + \frac{\pi(2k - 1)}{m} \leq \theta < z_i + \frac{\pi(2k + 1)}{m}  \\ 
\vdots & \\
b_i e^{-j2\pi}, &  z_i + \frac{\pi(2m - 1)}{m} \leq \theta < 2\pi. \\
\end{cases}
\end{equation}

Let 
\[
f(\theta) = \{ \hat{d}_i(\theta), i \in D \}
\]
be a function mapping the interval $[0, 2\pi)$ to a sequence of phase shift keyed symbols indexed by the elements of $D$.  Observe that $f(\theta)$ is piecewise continuous.  The subintervals of $[0, 2\pi)$ over which $f(\theta)$ remains contant are determined by the values of $\{z_i, i \in D\}$.  Let
\[
S = \{ f(\theta) \mid \theta \in [0, 2 \pi) \}
\]
be the set of all sequences $f(\theta)$ as $\theta$ varies from $0$ to $2\pi$.  If $\hat{\theta}$ is the least squares estimator the phase then $S$ contains the sequence $\{ \hat{d}_i(\hat{\theta}), i \in D \}$ corresponding to the least squares estimator of the data symbols, i.e., $S$ contains the minimiser of~\eqref{eq:SSdatasymbols}.  Observe from~\eqref{eq:uicombos} that there are atmost $\sum_{m\in G}(m+1)\abs{D_m} = O(L)$ sequences in $S$, because there are $m+1$ sequences for each $i \in D_m$.

The sequences in $S$ can be enumerated as follows.  Let 
\[
T = \bigcup_{m\in G} \bigcup_{ i \in D_m }\bigcup_{k=1}^{m}\left\{ (z_i + \tfrac{\pi(2k - 1)}{m}, i, m) \right\}
\]
be a set of triples with first element a real number, second element from $D$ and third element from $G$.  Let 
\[
H = \abs{T} = \sum_{m\in G}(m+1)\abs{D_m} = O(L)
\]  
denote the number of elements in $T$.  Let $t_1,\dots,t_H$ be an enumeration of the triples from $T$ sorted in accending order of the first element in the triple, that is, if $t_{i} = (t_{i1},t_{i2},t_{i3})$ then 
\[
t_{i1} \leq t_{ki}
\]
whenever $i < k$.  Put $\sigma(i) = t_{i2}$ and $m(i) = t_{i3}$ so that $\sigma(1), \dots, \sigma(H)$ and $m(1),\dots,m(H)$ correspond respectively with the second and third elements from the triples $t_1,\dots,t_H$ 
%So $a_1,\dots,a_H$ is a sequence of real numbers, $b_1,\dots,b_H$ is a sequence of integers from $D$ and $c_1,\dots,c_H$ is a sequence

The first sequence in $S$ is 
\[
f_1 = f(0) = \{ \hat{d}_i(0), i \in D \} = \{ b_i, i \in D \}.
\]  
The next sequence $f_1$ is given by replacing the element $b_{\sigma(1)}$ in $f_0$ with $b_{\sigma(1)}e^{-j2\pi/m(1)}$.  Given a sequence $x$ we use $x e_{k}$ to denote $x$ with the $\sigma(k)$th element replaced by $x_{\sigma(k)} e^{-j2\pi/m(k)}$.  Using this notation,  
\[
f_2 = f_1 e_1.
\] 
The next sequence in $S$ is correspondingly 
\[
f_3 = f_1 e_{1} e_{2} = f_2 e_{2},
\]
and the $k$th sequence is
\begin{equation}\label{eq:fkrec}
f_{k+1} = f_{k} e_{k}.
\end{equation}
In this way, all $(M+1)\abs{D}$ sequences in $S$ can be recursively enumerated.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{k}^*,
\]
where $g_{k} e_{k}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/m(k)}$.  Now,
\[
Y_1 = B + \sum_{i \in D} g_{1i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_2 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/m(1)} - 1)g_{1\sigma(1)} + \sum_{i \in D} g_{1i} \\
&= Y_0 + (e^{j2\pi/m(1)} - 1)g_{1\sigma(1)},
\end{align*}
In general,
\[
Y_{k+1} = Y_k + (e^{j2\pi/m(k)} - 1) g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in a constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata},
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Psuedocode is given in Algorithm~\ref{alg:loglinear}.  BLERG

%Line~\ref{alg_sortindices} contains the function $\operatorname{sortindicies}$ that, given $z = \{z_i, i \in D\}$, returns the permutation $\sigma$ as described in~\eqref{eq:sigmasortind}.  The $\operatorname{sortindicies}$ function requires sorting $\abs{D}$ elements.  This requires $O(L \log L)$ operations.  The $\operatorname{sortindicies}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

\begin{algorithm}[t] \label{alg:loglinear}
\SetAlCapFnt{\small}
\SetAlTitleFnt{}
\caption{Mackenthun's algorithm with pilot symbols}
\DontPrintSemicolon
\KwIn{$\{y_i, i \in P \cup D \}$}
\For{$m \in G$ \nllabel{alg_loop_setup}}{
\For{$i \in D_m$ }{
$\phi = \angle{y_i}$ \;
$u = \round{\phi}_m $ \;
$g_i = y_i e^{-j u}$ \;
\For{$k = 1, \dots, m$ }{ $t_{mik} =  (z_i + \tfrac{\pi(2k + 1)}{m}, i, m)$} 
}
}
$\operatorname{sort}(t_1,\dots,t_H)$ \;
$Y = \sum_{i \in P} y_i p_i^* + \sum_{i \in D} g_i $ \nllabel{alg_Y}\;
$\hat{a} = \frac{1}{L} Y$ \;
$\hat{Q} = \frac{1}{L}\abs{Y}^2$ \nllabel{alg_Q} \;
\For{$k = 1,\dots,H$ \nllabel{alg_loop_search}}{
$Y = Y + (e^{j2\pi/m(k)} - 1) g_{\sigma(k)}$ \;
$g_{\sigma(k)} = e^{j2\pi/m(k)} g_{\sigma(k)} $\;
$Q = \frac{1}{L}\abs{Y}^2$\;
\If{$Q > \hat{Q}$}{
 	$\hat{Q} = Q$ \;
 	$\hat{a} =  \frac{1}{L} Y$ \;
 }
}
\Return{$\hat{a}$ \nllabel{alg_return}}
\end{algorithm}

@@

% \begin{figure}[p]
% 	\centering
% 		\includegraphics[width=\linewidth]{code/data/plotM2-2.mps}
% 		\caption{Phase error versus SNR for BPSK with $L=4096$.}
% 		\label{fig:plotphaseBPSK}
% \end{figure}

>>>>

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-phase shift keying, and consider specifically the case where a number of different constellations sizes $M$ are used simultaneously.  This scenario is common in practical communications systems employing what is called \emph{unequal error proection}.  We describe an algorithm to compute the least squares estimators of carrier phase and amplitude that requires $O(L \log L)$ arithmetic operations, where $L$ is the number of recieved symbols.   
\end{abstract}
\begin{IEEEkeywords}
Coherent detection, noncoherent detection, phase shift keying, asymptotic statistics
\end{IEEEkeywords}

@@

In passband communication systems the transmitted signal typically undergoes time offset (delay), phase shift and attenuation (amplitude change).  These effects must be compensated for at the receiver. In this paper we assume that the time offset has been previously handled, and we focus on estimating the phase shift and attenuation.  We consider signalling constellations that have symbols evenly distributed on the complex unit circle such as binary phase shift keying (BPSK), quaternary phase shift keying (QPSK) and $M$-ary phase shift keying ($M$-PSK).  In this case, the transmitted symbols take the form,
\[
s_i = e^{j u_i},
\]
where $j = \sqrt{-1}$ and $u_i$ is from the set $\{0, \tfrac{2\pi}{M}, \dots, \tfrac{2\pi(M-1)}{M}\}$ and $M \geq 1$ is the size of the constellation.  We assume that some of the transmitted symbols are \emph{pilot symbols} known to the receiver and the remainder are information carrying \emph{data symbols} with phase that is unknown to the receiver.  So,
\[
s_i = \begin{cases}
p_i & i \in P \\
d_i & i \in D,
\end{cases}
\]
where $P$ is the set of indices describing the position of the pilot symbols $p_i$, and $D$ is a set of indices describing the position of the data symbols $d_i$.  The sets $P$ and $D$ are disjoint, i.e. $P \cap D = \emptyset$, and $L = \abs{P \cup D}$ is the total number of symbols transmitted.  The data symbols are further separated into subsets according to the size of the constellation used to modulate each symbol.  Let $D_2,D_3,\dots$ denote a partition of the indices in $D$ such that $D_2$ is the set of indices corresponding to data symbols modulated with $2$-PSK (i.e BSPK), and $D_3$ is the set of symbols modulated with $3$-PSK, and $D_4$ is the set of symbols modulated with $4$-PSK (i.e. QPSK) and so on.  Let $G$ be the set of integers for which $D_m$ is not empty, i.e. $D_m \neq \emptyset$ whenever $m \in G$.  We have 
\[
D = \bigcup_{m=2}^{\infty}D_m = \bigcup_{m\in G} D_m.
\]
In practical settings the number of different constellations used $\abs{G}$ is small.   A practical example might be that BPSK symbols are used for important symbols and QPSK symbols are used for less important symbols.  In this example $G = \{2,4\}$ and $\abs{G} = 2$.

@@

The existing literature~\cite{Mackenthun1994,Cowley_ref_sym_carr_1998,ViterbiViterbi_phase_est_1983,Sweldens2001,Wilson1989,Makrakis1990,Liu1991} mostly considers what is called \emph{noncoherent detection} where no pilot symbols exist ($P = \emptyset$ where $\emptyset$ is the empty set) and where only one constellation size is used, i.e., $G = \{M\}$ contains precisely one element.  In the noncoherent setting \emph{differential encoding} is often used, and for this reason the estimation problem has been called \emph{multiple symbol differential detection}.  A popular approach is the so called \emph{non-data aided}, sometimes also called \emph{non-decision directed}, estimator based on the paper of Viterbi and Viterbi~\cite{ViterbiViterbi_phase_est_1983}.  The idea is to `strip' the modulation from the recieved signal by taking $y_i / \abs{y_i}$ to the power of $M$.  A function $F: \reals \mapsto \reals$ is chosen and the estimator of the carrier phase $\theta_0$ is taken to be $\tfrac{1}{M}\angle{A}$ where $\angle$ denotes the complex argument and
\[
A = \frac{1}{L}\sum_{i \in P \cup D} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^M.
\]
Various choices for $F$ are suggested in~\cite{ViterbiViterbi_phase_est_1983} and a statistical analysis is presented.  A caveat is that it is not obvious how pilot symbols or data symbols using mulitple constellations should be included in this estimator.  % One approach is to compute
% \[
% A = \frac{1}{L} \sum_{i \in P} F(\abs{y_i}) \tfrac{y_i}{\abs{y_i}} p_i^* +  \frac{1}{L} \sum_{m \in G} \sum_{i \in D_m} F(\abs{y_i}) \big(\tfrac{y_i}{\abs{y_i}}\big)^m
% \]
% and choose the estimate of carrier phase to be $\tfrac{1}{K}\angle{A}$ where
% \[
% K = \begin{cases}
% 1 & \abs{P} > 0 \\
% \min(G) & \text{otherwise},
% \end{cases}
% \]
% is equal to $1$ when pilot symbols exists, and is otherwise equal to the size of the smallest constellation $\min(G)$.  The simulations we present in Section~\ref{sec:simulations} suggest that the accuracy of this estimator is poor when compared to the least squares estimator.  %This problem does not occur with the least square estimator.
This problem does not occur with the least squares estimator.

An important paper is by Mackenthun~\cite{Mackenthun1994} who described an algorithm to compute the least squares estimator requiring only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.  Both Mackenthun and Swelden's considered only the noncoherent setting and where only one constellation size is used, but we show in Section~\ref{sec:least-squar-estim}~that Mackenthun's algorithm can be modified to include pilot symbols and multiple constellation sizes. %Our model includes the noncoherent case by setting the number of pilot symbols to zero, that is, putting $P = \emptyset$.

@@

Define the sum of squares function
\begin{align}
SS(a, &\{d_i, i \in D\}) = \sum_{i \in P \cup D} \abs{ y_i - a s_i }^2 \nonumber \\
&= \sum_{i \in P \cup D} \abs{y_i}^2 - a s_i y_i^* - a^* s_i^* y_i + aa^*, \label{eq:SS}
\end{align}
where $*$ denotes the complex congugate.  Fixing the data symbols $\{d_i, i \in D\}$, differentiating with respect to $a^*$, and setting the resulting equation to zero we find the least squares estimator of $a_0$ as a function of $\{d_i, i \in D\}$,
\begin{equation}\label{eq:hata}
\hat{a}(\{d_i, i \in D\}) = \frac{1}{L} \sum_{i \in P \cup D} y_i s_i^* = \frac{1}{L} Y
\end{equation}
where $L = \abs{P \cup D}$ is the total number of symbols transmitted, and to simplify our notation we have put 
\[
Y = \sum_{i \in P \cup D} y_i s_i^* = \sum_{i \in P } y_i p_i^* + \sum_{i \in D } y_i d_i^*.
\]  
Note that $Y$ is a function of the unknown data symbols $\{ d_i, i \in D\}$ and we could write $Y(\{ d_i, i \in D\})$, but have chosen to surpress the argument $(\{ d_i, i \in D\})$ for notational clarity.  Substituting $\frac{1}{L}Y$ for $a$ into~\eqref{eq:SS} we obtain the $SS$ conditioned on minimisation with respect to $a$,
\begin{equation}\label{eq:SSdatasymbols}
SS(\{d_i, i \in D\}) = A - \frac{1}{L}\abs{Y}^2,
\end{equation}
where $A = \sum_{i \in P \cup D}\abs{y_i}^2$ is a constant.  Observe that given candidate values for the data symbols, we can compute the corresponding $SS(\{d_i, i \in D\})$ in $O(L)$ arithmetic operations.  Let 
\[
H = \sum_{m\in G}m\abs{D_m} = O(L).
\]
It turns out that there are atmost $H$ candidate values of the least squares estimator of the data symbols~\cite{Sweldens2001,Mackenthun1994}.  %When the number of data symbols is not small, this set is substantially smaller than the entire set of possible transmitted symbols, which is of size $M^{\abs{D}}$.

@@

We only require to consider $\theta$ in the interval $[0, 2\pi)$.  Consider how $\hat{d}_i(\theta)$ changes as $\theta$ varies from $0$ to $2\pi$.  Let $b_i = \hat{d}_i(0)$ and let 
\[
z_i = \angle{y_i} - \hat{u}_i(0) = \angle{y_i} - \round{\angle{y_i}}_{m}.
\]
Then,
\begin{equation}\label{eq:uicombos}
\hat{d}_i(\theta) = 
\begin{cases}
b_i, &  0 \leq \theta < z_i + \frac{\pi}{m} \\
b_i e^{-j2\pi/m}, & z_i + \frac{\pi}{m} \leq \theta < z_i + \frac{3\pi}{m} \\ 
\vdots & \\
b_i e^{-j2\pi k /m}, & z_i + \frac{\pi(2k - 1)}{m} \leq \theta < z_i + \frac{\pi(2k + 1)}{m}  \\ 
\vdots & \\
b_i e^{-j2\pi} = b_i, &  z_i + \frac{\pi(2m - 1)}{m} \leq \theta < 2\pi. \\
\end{cases}
\end{equation}

Let 
\[
f(\theta) = \{ \hat{d}_i(\theta), i \in D \}
\]
be a function mapping the interval $[0, 2\pi)$ to a sequence of phase shift keyed symbols indexed by the elements of $D$.  Observe that $f(\theta)$ is piecewise continuous.  The subintervals of $[0, 2\pi)$ over which $f(\theta)$ remains contant are determined by the values of $\{z_i, i \in D\}$.  Let
\[
S = \{ f(\theta) \mid \theta \in [0, 2 \pi) \}
\]
be the set of all sequences $f(\theta)$ as $\theta$ varies from $0$ to $2\pi$.  If $\hat{\theta}$ is the least squares estimator the phase then $S$ contains the sequence $\{ \hat{d}_i(\hat{\theta}), i \in D \}$ corresponding to the least squares estimator of the data symbols, i.e., $S$ contains the minimiser of~\eqref{eq:SSdatasymbols}.  Observe from~\eqref{eq:uicombos} that there are atmost $H = O(L)$ sequences in $S$, because there are $m$ distinct values of $\hat{d}_i(\theta)$ for each $i \in D_m$.

The sequences in $S$ can be enumerated as follows.  Let 
\[
T = \bigcup_{m\in G} \bigcup_{ i \in D_m }\bigcup_{k=1}^{m}\left\{ (z_i + \tfrac{\pi(2k - 1)}{m}, i, m) \right\}
\]
be a set of triples with first element a real number, second element from $D$ and third element from $G$.  The number of elements in $T$ is $H = \abs{T} = O(L)$.  Let $t_1,\dots,t_H$ be an enumeration of the triples from $T$ sorted in accending order of the first element in the triple, that is, if $t_{k} = (t_{k1},t_{k2},t_{k3})$ then $t_{i1} \leq t_{k1}$ whenever $i < k$.  Put 
\[
\sigma(k) = t_{k2} \qquad \text{and} \qquad m(k) = t_{k3}
\]
so that $\sigma(1), \dots, \sigma(H)$ and $m(1),\dots,m(H)$ correspond respectively with the second and third elements from the triples $t_1,\dots,t_H$.
%So $a_1,\dots,a_H$ is a sequence of real numbers, $b_1,\dots,b_H$ is a sequence of integers from $D$ and $c_1,\dots,c_H$ is a sequence

The first sequence in $S$ is 
\[
f_1 = f(0) = \{ \hat{d}_i(0), i \in D \} = \{ b_i, i \in D \}.
\]  
The next sequence $f_1$ is given by replacing the element $b_{\sigma(1)}$ in $f_0$ with $b_{\sigma(1)}e^{-j2\pi/m(1)}$.  Given a sequence $x$ we use $x e_{k}$ to denote $x$ with the $\sigma(k)$th element replaced by $x_{\sigma(k)} e^{-j2\pi/m(k)}$.  Using this notation,  
\[
f_2 = f_1 e_1.
\] 
The next sequence in $S$ is correspondingly 
\[
f_3 = f_1 e_{1} e_{2} = f_2 e_{2},
\]
and the $k$th sequence is
\begin{equation}\label{eq:fkrec}
f_{k+1} = f_{k} e_{k}.
\end{equation}
In this way, all sequences in $S$ can be recursively enumerated.

@@

Let,
\begin{equation}\label{eq:SSfk}
SS(f_k) = A - \frac{1}{L}\abs{Y_k}^2,
\end{equation}
where, 
\begin{align*}
Y_k = Y( f_k ) &= \sum_{i \in P} y_i p_i^*  + \sum_{i \in D} y_i f_{ki}^* \\
&= B + \sum_{i \in D}g_{ki},
\end{align*}
where $B = \sum_{i \in P} y_i p_i^*$ is a constant, independent of the data symbols, and $f_{ki}$ denotes the $i$th symbol in $f_k$, and for convenience, we put $g_{ki}  = y_i f_{ki}^*$.  Letting $g_{k}$ be the sequence $\{g_{ik}, i \in D\}$ we have, from~\eqref{eq:fkrec}, that $g_k$ satisfies the recursive equation
\[
g_{k+1} = g_{k} e_{k}^*,
\]
where $g_{k} e_{k}^*$ indicates the sequence $g_k$ with the $\sigma(k)$th element replaced by $g_{k \sigma(k)}e^{j2\pi/m(k)}$.  Now,
\[
Y_1 = B + \sum_{i \in D} g_{1i}
\] 
can be computed in $O(L)$ operations, and
\begin{align*}
Y_2 &= B + \sum_{i \in D} g_{1i} \\
&= B +  (e^{j2\pi/m(1)} - 1)g_{1\sigma(1)} + \sum_{i \in D} g_{1i} \\
&= Y_0 + (e^{j2\pi/m(1)} - 1)g_{1\sigma(1)},
\end{align*}
In general,
\[
Y_{k+1} = Y_k + (e^{j2\pi/m(k)} - 1) g_{k\sigma(k)}.
\]
So, each $Y_k$ can be computed from it predecessor $Y_{k-1}$ in a constant number of operations.  Given $Y_k$, the value of $SS(f_k)$ can be computed in a constant number of operations using~\eqref{eq:SSfk}.  Let $\hat{k} = \arg\min SS(f_k)$.  The least squares estimator of the complex amplitude is then computed according to~\eqref{eq:hata},
\begin{equation}\label{eq:ahatYhat}
\hat{a} = \frac{1}{L} Y_{\hat{k}}.
\end{equation}
Pseudocode is given in Algorithm~\ref{alg:loglinear}.   The function $\operatorname{sort}$ on line~\ref{alg_sortindices} return the triple $t_1,\dots,t_H$ so that the first element in the triples is in ascending order.   This requires $O(L \log L)$ operations.  The $\operatorname{sort}$ function is the primary bottleneck in this algorithm when $L$ is large.  The loops on lines~\ref{alg_loop_setup} and~\ref{alg_loop_search} and the operations on lines~\ref{alg_Y} to lines~\ref{alg_Q} all require $O(L)$ or less operations.  %In the next sections we will show how to the sortinces function can be avoided.  This leads to an algorithm that requires only $O(L)$ operations.

\begin{algorithm}[t] \label{alg:loglinear}
\SetAlCapFnt{\small}
\SetAlTitleFnt{}
\caption{Mackenthun's algorithm with pilot symbols}
\DontPrintSemicolon
\KwIn{$\{y_i, i \in P \cup D \}$}
$c = 1$ \;
\For{$m \in G$ \nllabel{alg_loop_setup}}{
\For{$i \in D_m$ }{
$\phi = \angle{y_i}$ \;
$u = \round{\phi}_m $ \;
$g_i = y_i e^{-j u}$ \;
$z  = \phi - u$ \;
\For{$k = 1, \dots, m$ }{ 
$t_{c} =  (z + \tfrac{\pi(2k - 1)}{m}, i, m)$ \;
$c = c + 1$ \;
} 
}
}
$\operatorname{sort}(t_1,\dots,t_H)$ \nllabel{alg_sortindices} \;
$Y = \sum_{i \in P} y_i p_i^* + \sum_{i \in D} g_i $ \nllabel{alg_Y}\;
$\hat{a} = \frac{1}{L} Y$ \;
$\hat{Q} = \frac{1}{L}\abs{Y}^2$ \nllabel{alg_Q} \;
\For{$k = 1,\dots,H$ \nllabel{alg_loop_search}}{
$Y = Y + (e^{j2\pi/m(k)} - 1) g_{\sigma(k)}$ \;
$g_{\sigma(k)} = e^{j2\pi/m(k)} g_{\sigma(k)} $\;
$Q = \frac{1}{L}\abs{Y}^2$\;
\If{$Q > \hat{Q}$}{
 	$\hat{Q} = Q$ \;
 	$\hat{a} =  \frac{1}{L} Y$ \;
 }
}
\Return{$\hat{a}$ \nllabel{alg_return}}
\end{algorithm}

@@

 \begin{figure}[p]
 	\centering
 		\includegraphics[width=\linewidth]{code/data/plot-2.mps}
 		\caption{Phase error versus SNR}
 		\label{fig:plotphaseBPSK}
 \end{figure}
